What We Eval in the Shadows

### 2021-04-17 paper/main.tex
Symbolic differentiation

The ability to capture current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allowing one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

@@

% TODO Pierre
For instance, base R packages \emph{stats} provides two operators for \emph{symbolic differentiation}, \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created, which contains expressions derived twice (with \c{DD}) and then typically in a list or a data frame.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}
% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd

@@

Finally, let us mention an anecdotal use of \eval as a mean to bypass some
checks performed by the \c{R CMD CHECK} tool. This tool performs a numer of
well-formedness rules before accepting a package to the CRAN repository. Many
checks are performed statically, by looking at syntactic constructs in the R
code. Using \eval allows one to bypass such checks and thus submit a package to
CRAN that would otherwise be refused. For example, the it checks that packages
do not call \c{unlockBinding} function.\footnote{It allows one to open and
modify loaded package name space.} Packages such as \c{data.table} circumvent
such restrictions by using \eval as the check is only performed statically.

>>>>

The ability to capture current call as a language object (\c{match.call()}),
extract the unevaluated expression from function arguments (\c{substitute()})
and \eval with first class environment makes the R toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \c{data.table}
package defines a DSL for data frame manipulation using the subset operator
\c{[}. This allows one to write compact queries such as \c{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \c{data.table} subset operation is
defined as \c{DT[i, j, k]}, where \c{j} expression is used to calculate on a
subset/reorder of rows from \c{i} grouped by \c{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \c{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

@@

Another instance, is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \c{D} and \c{deriv}. They support arithmetic operations and functions on real numbers such as \c{sin}, on expressions built with \c{expression()} or \c{parse}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \c{Dtensor} object is created: it contains expressions that are derived twice (with \c{DD}). To get a concrete value, it is evaluated in  dataframe \c{mu} where symbolic variables are bound to actual numbers.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

% See https://github.com/cran/MCMCglmm/blob/cfac9e67ec73a3db2142826faae35e5d2318da31/man/evalDtensor.Rd

@@

Finally, let us mention an anecdotal use of \eval as a mean to bypass some
checks performed by the \c{R CMD CHECK} tool allowing one to submit a package
to CRAN that would otherwise be refused. For example, the tool checks that
packages do not call \c{unlockBinding} function.\footnote{It allows one to open
and modify loaded package name space.} Packages such as \c{data.table}
circumvent such restrictions by using \eval as the check is only performed
statically.
